---
title: "Data Viz Proj"
author: "Shreyas Meher"
date: "11/27/2021"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    social: menu
    source_code: embed
---

```{r setup, include=FALSE}

library(rgdal) # plot map
library(leaflet) # plot map
library(dplyr) # data wrangling
library(tidyr) # data wrangling
library(ggplot2) # plot
library(plotly) # interactive plot
library(shiny) # shiny
library(shinydashboard) # shiny dashboard
library(ggthemes) # themes for ggplot2

knitr::opts_chunk$set(echo = TRUE)
```

Ecological Footprint according to GDP {data-orientation=columns}
====================================================

Sidebar {.sidebar}
----------------------------------------------------
### Ecological Footprint ~ GDP

*To find out the relationship between the GDP of the country and the Ecological footprint, I am using a simple linear plot to estimate the relationship.
*With the output, we can see that as GDP increases, the Ecological footprint increases correspondingly. The data used is from the Global Footprint Network website (https://www.footprintnetwork.org/) and the dataset is freely accessible to all. 
*I created a unique dataset gleaning from a few variables from the GFN along with World Bank data. 


```{r}
# data loading

footprint <- read.csv("countries.csv")

library(rgdal) # plot map
library(leaflet) # plot map
library(dplyr) # data wrangling
library(tidyr) # data wrangling
library(ggplot2) # plot
library(plotly) # interactive plot
library(shiny) # shiny
library(shinydashboard) # shiny dashboard
library(ggthemes) # themes for ggplot2

# GDP and Ecological footprint

ggplot(footprint, aes(x=HDI, y=Total.Ecological.Footprint))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(y = "Total Ecological Footprint in global hectares",
       x = 'GDP',
       title = "Ecological Footprint according to GDP",
       caption = 'Source: Global Ecoogical Footprint, Global Footprint Network')
```


